### 2021-06-04 papers/thesis/full/ch2-spherical-harmonics/spherical-harmonics.tex
Further edits to help make sure the thesis flows ok and legacy paper specific language is removed from ch3 and ch4.

\chapter{Spherical harmonics as orthogonal polynomials}

@@

In this chapter, we demonstrate how one can write the spherical harmonics as polynomials in $(x, y, z)$, and how the relationships between them can be used to obtain sparse \enquote{Jacobi operators} for multiplication by $x,y,z$. We demonstrate how these in turn lead to an algorithm for evaluating the spherical harmonics, an efficient algorithm for evaluating functions when expanded in the SH OP basis, and sparse \enquote{banded-block-banded} differential operator matrices. The techniques used here are done so with the aim of applying them to other OP families of other domains (and we do so in later chapters). We will finally demonstrate a proof-of-concept via the examples of the heat equation and the linearised shallow water equations on the unit sphere.

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. It is useful to be able to transform between these cartesian coordinates and the spherical coordinates ($\varphi, \theta$). Throughout this document we will use the convention that the spherical coordinate angles be defined by
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi.
}
where $\rho(z) := \sqrt{1-z^2}$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[14.30.1]{DLMF}:
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases} \label{eqn:clmdef}
}
which can be simplified to
\bseqn{
	\clm = \Bigg({(2l+1) \: (l+m)! \: (l-m)! \over \pi }\Bigg)^\half \: {1 \over 2^{\abs{m} + 1} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases}.
}
The spherical harmonics as defined are then orthonormal with respect to the inner product
\bseqn{
	&\int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	& \quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	& \quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x,y,z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x,y,z$ for any $m \in \Z$. Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in $\R^3$.

@@

We can use the Clenshaw algorithm to evaluate this function at a given point $(x,y,z)$ on $\Omega$ \cite{press2007numerical}. The \textit{Clenshaw algorithm} is then as follows:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \bold{0}, \: \vec{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \fvec_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \vec{\xi}_{0}^T \SHvec_0(x,y,z) \equiv \xi_0 Y_0.
}

@@

%%% TANGENT BUNDLE ----- VECTOR SHs
\section{Vector spherical harmonics as orthogonal vectors in three variables}\label{Section:VSHs}

@@

The framework we have outlined for the spherical harmonics and the vector spherical harmonics allows us to easily and explicitly derive sparse operator matrices for partial differential operators such as the spherical gradient, divergence and Laplacian, as well as other key operators useful in examples such as the linearised shallow water equations. These operators will take coefficients of a function's expansion in either the SSH or VSH basis, to coefficients in either basis.

@@

\begin{definition}\label{def:SHoperators}
	We define the operator matrices $\calD, \tilde \calD, \calG, \tilde \calG \calL$ according to:
\bseqn{
	\divergence \uvec(x,y,z) &= \SHvec(x,y,z)^\top \: \calD \: \uvecc, \\
	\divergence \uvec(x,y,z) &= \SHvec(x,y,z)^\top \: \calE_s \: \big(\tilde \calD \: \uvec^c), \\
	\gradS f(x,y,z) &= \VSHvec(x,y,z)^\top \: \calG \: \fvecc, \\
	\gradS f(x,y,z) &= \VSHvecfull(x,y,z)^\top \: \tilde \calG \: \fveccexp, \\
	\DeltaS f(x,y,z) &= \SHvec(x,y,z)^\top \: \calL \: \fvecc.
}
\end{definition}
\begin{remark}
	The operators $\tilde \calD, \tilde \calG$ would be used when vectors of the resulting length are required for the problem you are wishing to solve.
\end{remark}

>>>>

\chapter{Spherical harmonics as orthogonal polynomials}\label{CHAPTER:sphericalharmonics}

@@

It is useful to be able to transform between these cartesian coordinates and the spherical coordinates ($\varphi, \theta$). Throughout this thesis we will use the convention that the spherical coordinate angles be defined by
\bseqn{
	x &= \sinphi \costheta = \rho(z) \costheta \\
	y &= \sinphi \sintheta = \rho(z) \sintheta \\
	z &= \cosphi
}
where $\rho(z) := \sqrt{1-z^2}$.

In this chapter, we demonstrate how one can write the spherical harmonics as polynomials in $(x, y, z)$, and how the relationships between them can be used to obtain sparse \enquote{Jacobi operators} for multiplication by $x,y,z$. We demonstrate how these in turn lead to an algorithm for evaluating the spherical harmonics, an efficient algorithm for evaluating functions when expanded in the SH OP basis, and sparse \enquote{banded-block-banded} differential operator matrices. The techniques used here are done so with the aim of applying them to other OP families of other domains (and we do so in later chapters). We will finally demonstrate a proof-of-concept via the examples of the heat equation and the linearised shallow water equations on the unit sphere. These examples consider partial differential operators involving the spherical Laplacian (the Laplace--Beltrami operator): written in spherical coordinates this is
\bseqnnumber{
	\DeltaS &= {1 \over \sinphi} \ppphi \Big( \sinphi \ppphi \Big) + {1 \over \sin^2 \varphi} \ppthetatwo = {1 \over \rho} \ppphi \Big( \rho \ppphi \Big) + {1 \over \rho^2} \ppthetatwo \label{eqn:laplacebeltrami}
}
i.e. $\DeltaS f(\xvec) = \Delta f({\xvec \over \norm{\xvec}})$ for some function $f$ where $\xvec := (x,y,z) \in \R^3$.

@@

On that note, we can now write down the spherical harmonics. Let $(x,y,z) \in \Omega$. We will use the standard definition -- that is, the spherical harmonics, orthonormal on the unit sphere, are \cite[14.30.1]{DLMF}:
\bseqnnumber{
	\Ylmfull &:= \Bigg({(2l+1) \: (l-m)! \over 4 \pi \: (l+m)!}\Bigg)^\frac{1}{2} \: \eimtheta \Plm (\cosphi) \nonumber \\
	&= \clm (1 - (\cosphi)^2)^\frac{|m|}{2} \eimtheta P^{(|m|,|m|)}_{l-|m|}(\cosphi) \nonumber \\
	&= \clm P^{(|m|,|m|)}_{l-|m|}(z) \: \rho(z)^{|m|} \: \eimtheta \label{eqn:shdef}
}
for $0 \le |m| \le l, \, l \in \No$, where
\bseqnnumber{
	\clm := \Bigg({(2l+1) \: (l-m)! \over 4\pi \: (l+m)!}\Bigg)^\half \: {(l+m)! \over 2^{\abs{m}} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases} \label{eqn:clmdef}
}
which can be simplified to
\bseqn{
	\clm = \Bigg({(2l+1) \: (l+m)! \: (l-m)! \over \pi }\Bigg)^\half \: {1 \over 2^{\abs{m} + 1} \; l!} \:
		\begin{cases} 
			(-1)^m &\text{if } m \ge 0 \\
			1 &\text{if } m < 0
		\end{cases}.
}
The spherical harmonics as defined are then orthonormal with respect to the inner product
\bseqn{
	&\int_0^{2\pi} \int_0^\pi \Ylmfull \: {Y_{l'}^{m'}(\varphi, \theta)}^{*} \: \sinphi \: \D \varphi \: \D \theta \\
	& \quad = 2\pi \: \delta_{m, m'} \: \clm \: c_{l'}^{m'} \: \int_{-1}^1 P^{(|m|,|m|)}_{l-|m|}(z) \: P^{(|m|,|m|)}_{l'-|m|}(z) \: \rho(z)^{2|m|} \: \D z \\
	& \quad = \delta_{l, l'} \: \delta_{m, m'}
}
where $\alpha^*$ denotes the complex conjugate of $\alpha \in \C$. Note how we can express the spherical harmonics $\Ylm$ in terms of $x,y,z$ instead of $\varphi, \theta$ by noting that $\rho(z)^{|m|} \eimtheta$ can be expressed in terms of $x,y,z$ for any $m \in \Z$. Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the vanishing ideal associated by the zero set of $x^2 + y^2 + z^2 - 1$ in $\R^3$.

@@

A multivariate analogue of Clenshaw's recursive algorithm for evaluation of a function expanded in an OP basis has been established for methods on other domains in 2D such as the triangle \cite{olver2019triangle}, and we proceed similarly here. We can use the Clenshaw algorithm to evaluate this function at a given point $(x,y,z)$ on $\Omega$ \cite{press2007numerical}. The multidimensional \textit{Clenshaw algorithm} is as follows:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \bold{0}, \: \vec{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:0 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^T = \fvec_n^T - \vec{\xi}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \vec{\xi}_{n+2}^T D^T_{n+1}C_{n+1} \\
	\quad &\text{3) } \text{Output: } f(x,y,z) \approx \vec{\xi}_{0}^T \SHvec_0(x,y,z) \equiv \xi_0 Y_0.
}

@@

%%% TANGENT BUNDLE ----- VECTOR SHs
\section{Vector spherical harmonics as orthogonal vectors in three variables}\label{section:VSHs}

@@

The framework we have outlined for the spherical harmonics and the vector spherical harmonics allows us to easily and explicitly derive sparse operator matrices for partial differential operators such as the spherical gradient, divergence and Laplace--Beltrami, as well as other key operators useful in examples such as the linearised shallow water equations. These operators will take coefficients of a function's expansion in either the SSH or VSH basis, to coefficients in either basis.

@@

\begin{definition}\label{def:SHoperators}
	We define the operator matrices $\calD, \tilde \calD, \calG, \tilde \calG, \calL$ according to:
\bseqn{
	\divergence \uvec(x,y,z) &= \SHvec(x,y,z)^\top \: \calD \: \uvecc, \\
	\divergence \uvec(x,y,z) &= \SHvec(x,y,z)^\top \: \calE_s \: \big(\tilde \calD \: \uvec^c), \\
	\gradS f(x,y,z) &= \VSHvec(x,y,z)^\top \: \calG \: \fvecc, \\
	\gradS f(x,y,z) &= \VSHvecfull(x,y,z)^\top \: \tilde \calG \: \fveccexp, \\
	\DeltaS f(x,y,z) &= \SHvec(x,y,z)^\top \: \calL \: \fvecc.
}
\end{definition}
\begin{remark}
	The operators $\tilde \calD, \tilde \calG$ would be used when vectors of the resulting length are required for the problem you are wishing to solve.
\end{remark}

For clarity, we note that $\DeltaS$ is defined in \bsrefeqn{eqn:laplacebeltrami}, and the spherical gradient and divergence operators are given by
\bseqn{
	 \gradS &= \ppx{\varphi} + {1 \over \sinphi} \ppx{\theta}, \\
	 \divergence \uvec &= {1 \over \sinphi} \ppx{\varphi}\big( \sinphi u_\varphi \big) + {1 \over \sinphi} \ppx{\theta} u_\theta,
}
where $\uvec \equiv \phivec \: u_\varphi + \thetavec \: u_\theta$.
